---
title: "Hw03"
author: "Vetle Birkeland Huglen"
date: "1 10 2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
```

Firstly we look at the distribution of gdpPercap by continent. We can make a table showing som key values, and also violin plots, which are really nice for summarising a distribution.

```{r}
gdp_overview <- gapminder %>%
  group_by(continent) %>%
  summarize(max_gdpPercap = max(gdpPercap),
              min_gdpPercap = min(gdpPercap),
              gdpSpread = max_gdpPercap-min_gdpPercap)
  
knitr::kable(gdp_overview)

gapminder %>% 
  group_by(continent) %>%
  ggplot(aes(continent, gdpPercap, fill = continent)) +
  geom_violin()
```
The violin plots are really dragged out. This case may be better visualised by using a log function on the values:

```{r}
gapminder %>% 
  group_by(continent) %>%
  ggplot(aes(continent, log(gdpPercap), fill = continent)) +
  geom_violin()
```
Now lets look at a way of removing outliers from a statistic. We can compare the so called trimmed mean to the regular mean, to see if there is a big difference.

```{r}
gapminder %>%
  group_by(year) %>%
  summarize(trimMeanLifeExp = mean(lifeExp, trim=0.25),
            regMeanLifeExp = mean(lifeExp)) %>%
  ggplot() +
  geom_line(aes(year, trimMeanLifeExp), color = 'red') +
  geom_line(aes(year, regMeanLifeExp), color = 'blue')
```
Here follows a simple plot of how the life expectancy has changed over the years in each continent. I really like the geom_smooth function, since it also show the variance of the data.

```{r}
gapminder %>%
  group_by(continent) %>%
  ggplot(aes(year, lifeExp)) +
  geom_smooth(aes(colour = continent))
  
  #geom_bar(stat = "identity", aes(fill = continent))
```
And now to the fourth task. I found this one really hard, but I dont really know why. Now that its working, it seems to be in a very intuitive way. The table and graph shows the number and percentage of countries below the mean life expectancy in their respective continents, for each year. 

```{r}
countries_under_mean <- gapminder %>%
  group_by(year, continent) %>%
  summarise(mean = mean(lifeExp),
            n_under_mean = sum(lifeExp < mean),
            percent_under_mean_lifeExp = n_under_mean/n())

knitr::kable(countries_under_mean)

countries_under_mean %>%
  ggplot(aes(year, percent_under_mean_lifeExp)) +
  geom_line(aes(colour = continent))
```

